---
title: "Setting up R and R Studio"
author: "Robin Donatello"
date: "Math 199 - Spring 2016"
output: pdf_document
---

##  Goals of this lesson
This lesson will provide information on how to

1. Install R  -- _The programming language_
2. Install R Studio -- _A fantastic interface for coding in R_
3. Basics of Navigating R Studio
4. Set up preferences -- _To avoid common mistakes and pitfalls_
5. Install Packages -- _Adds functionality to R_


## Installing R & R Studio

Download the latest versions of R and R Studio for your flavor of operating system.

* [R](http://cran.r-project.org/)
* [R Studio](http://www.rstudio.com/products/rstudio/download/)

If you are having difficulty installing these programs here are links to helpful YouTube video on installing both R and R Studio. 

- [Mac Instructions](http://www.youtube.com/watch?v=uxuuWXU-7UQ)
- [Windows Instructions](http://www.youtube.com/watch?v=Ohnk9hcxf9M)


## Layout of R Studio

- The panel in the upper right contains your *workspace* (Environment) as well as a 
history of the commands that you've previously entered.  
- Any plots that you generate will show up in the panel in the lower right corner.
- There are multiple tabs down there such as your file explorer, and the help files. 
- The panel on the left is the *console* and it's where the action happens.

## The Console

- Everytime you launch RStudio, it will have the same text at the top of the 
console telling you the version of R that you're running.  
- Below that information is the *prompt*, `>` . As its name suggests, this prompt is really a request, a request for a command. 
- Initially, interacting with R is all about typing commands and interpreting the output. 
- These commands and their syntax have evolved over decades (literally) and now provide what many users feel is a fairly natural way to access data and organize, describe, and invoke statistical computations.


## Initial Setup
To avoid some of the most common errors when learning R let's set some preferences in RStudio. 

1. On your computer, create a folder called `RBootcamp`. Store all files associated with this class here.
2. In RStudio the file menu go to _Tools_ then _Global Options_. 
3. Select your `RBootcamp` folder as your default working directory by clicking [Browse.] and navigating to your `RBootcamp` folder. 
4. Uncheck "Restore .RData into workspace at startup" and "Always save history"
5. Where it says 'Save workspace to .RData on exit:" Select 'Never'
6. Check" "Remove duplicate entries in history"
7. Click _apply_ then _ok_. 

## Installing Packages 
* Packages are sets of predefined code, functions and data sets. They provide additional functionality and goodies beyond what is included when you download Base R. 

* <div class="red">Packages are only installed once per computer ever</div> 

* During this course you will be asked to complete several interactive tutorials that use code found in the `swirl` package. 

* To install the Swirl package type the following command at the prompt `>` and hit enter to submit the command.
```{r, eval=FALSE}
install.packages("swirl")
```

* Install the following packages, remembering `R` is case sensitive: `ggplot2`, `mail`, `dplyr`, `knitr`, `nycflights13` and `hflights`. 


```{r, eval=FALSE}
install.packages(c("ggplot2","mail","dplyr","knitr","nycflights13","hflights"))
```

## Downloading Swirl lessons

* The `swirl` package and it's fantastic tutorials were created by the folks at [http://swirlstats.com/](http://swirlstats.com/). All of their work is free to use and modify under the GPLv3 copyright license. 

* I have spliced and diced most of their prebuilt lessons to fit this class. My thanks to them for doing all the hard work. 

* Download the [zipped file](http://norcalbiostat.github.io/R-Bootcamp/swirl_lessons.zip) and place it into your `RBootcamp` folder. Do not unzip it. 
    - **MAC USERS - RIGHT CLICK THE LINK AND SELECT SAVE LINK AS**

* Leave a window open to this folder. You will need to know the path to this zip file next. 

## Finding your path on a Mac

1. Select the file or folder in the OS X Finder, then hit Command+i to summon Get Info
2. Get Info can also be accessed by the control-click and right-click menus.
3. Click and drag alongside "Where" to select the path, then hit Command+C to copy the full path to the clipboard 
4. Paste this into `R` for your path. 
5. Make sure all slashes are forward slashes `/`.

Mac Reference link: <http://bit.ly/1NxF9eu>


## Finding your path on Windows

1. Open Windows Explorer and browse to the folder or file of your choice.
2. The path to the folder/file will be indicated in the Address bar of Windows Explorer.
3. To copy it, simply right-click on the Address bar and select _Copy address as text_.
4. Paste this into `R` for your path. 
5. Make sure all slashes are forward slashes `/`.


Windows Reference link: <http://bit.ly/1NxFeir>


## Installing Swirl lessons
The functions used to install the custom lessons are contained within the `swirl` package. In the R console window type: 
```{r, eval=FALSE}
library(swirl)
install_course_zip("YOUR PATH HERE/swirl_lessons.zip")
```
* You must change `YOUR PATH HERE` to the path to where your RBootcamp folder is located. 
* Only change the `PATH`. 
* Be sure all slashes are forward `/`
* You will *_FREQUENTLY_* have to insert this `PATH`, so if you have any questions **NOW** is the time to ask!


## Using Swirl

* You'll learn how to use Swirl in [Lab 1](https://raw.githubusercontent.com/norcalbiostat/R-Bootcamp/gh-pages/labs/Lab1_-_Getting_Started_in_R.pdf)!
* It is **highly** recomended that you save a copy of every lab in your `RBootcamp` folder for later reference. 
* Don't forget to use the [Google Group](https://groups.google.com/forum/#!forum/csuc_stat) to ask questions if you get stuck.
    - Tag your posts with #Rbootcamp so it's easy to find!

