---
title: "Lab 4"
output: html_document
---

1.1 Factors

```{r}
table(iris$Species)
class(iris$Species)
a <- as.character(iris$Species)
class(a)
```


1.2 Lists

```{r}
my_vec <- sample(1:10, 5)
my_mat <- matrix(rnorm(1:10), nrow = 5)
my_df <- iris[1:5, ]

my_list <- list(vec = my_vec, mat = my_mat, df = my_df)
my_list

str(my_list)
```


1.3 Functions and their arguments

?mean

1.3.1 Argument Ordering

```{r}
y <- sample(1:100, 20)
y[sample(1:length(y), 5)] <- NA
y

mean(y)
mean(y, na.rm = TRUE)

mean(y, TRUE)

mean(na.rm = TRUE, x = y, trim = .1)
```

2. Practice

Use the NCbirths data set for the following questions.

```{r import}
NCbirths <- read.csv("NCbirths.csv", header = TRUE)
```

1. Calculate the mean age of the mothers (mage) in the sample.
  
```{r mage_mean}
mean(NCbirths$mage)
```

2. Calculate the trimmed mean mother's age after eliminating the outer 10% of ages.

This can be intrepretted two ways one is 10% is 5% from each side or 10% is 10% from each side.

```{r trimmed_mage_mean}
mean(NCbirths$mage, trim = 0.05)

mean(NCbirths$mage, trim = 0.1)
```

3. Calculate the mean father's age (fage).

```{r fage_mean}
mean(NCbirths$fage, na.rm = TRUE)
```

4. Do mothers who smoke give birth to smaller babies?  (Hint: Use subset to create subsets of the data set by smoking habit status then find the mean weight for each subset.)

```{r small_babies}
nonsmoke <- subset(NCbirths, habit == "nonsmoker")
smoke <- subset(NCbirths, habit == "smoker")
mean(nonsmoke$weight)
mean(smoke$weight)
```

In this data set on average the mothers who smoke give birth to smaller babies.

5. Do more non white mothers smoke?  Use table() to find this answer and don't forget to check for missing values.

```{r mother_smoke}
no_NA_NCbirths <- subset(NCbirths, !is.na(whitemom) & !is.na(habit))
save_table <- table(no_NA_NCbirths$habit, no_NA_NCbirths$whitemom)
save_table
prop_save_table <- prop.table(save_table, margin = 2)
prop_save_table
```

I don't think we have introduced the `prop.table` function by this lesson, but it shows that when looked at as proportions there is a higher propensity for white mothers in this data set to smoke than non white mothers.





