---
title: "Lab 1 Answers"
output: html_document
---

1.1 Basic Building Blocks
```{r, eval = FALSE}
library(swirl)
swirl()
```

1.1.1 Practice

1. Use sample(0:100, size=5) to create a vector of 5 random numbers between 0 and 100, inclusive.  Store this result as a variable x.

```{r}
x <- sample(0:100, size = 5)
```

2. Calculate the following (x-5)^2 / sqrt(50-1)

```{r}
(x-5)^2 / sqrt(50-1)
```


1.2 Sequences of Numbers

```{r, eval = FALSE}
swirl()
```

1.2.1 Practice

1. List the even numbers from 1 to 12.

```{r}
seq(2, 12, by = 2)
```

2. List the numbers 10 to 1 in descending order.

```{r}
10:1
seq(10,1)
```

3. Create this sequence using R: ("A", "B", "C", "A", "B", "C")

```{r}
rep(c("A", "B", "C"), 2)
```

4. Create this sequence using R: ("A", "A", "B", "B", "C", "C")

```{r}
rep(c("A", "B", "C"), each = 2)
```


1.3 Logic

```{r}
swirl()
```

1.3.1 Practice

1. Evaluate ((111>=111) | TRUE) & ((4 + 1) != 5)

```{r}
((111>=111) | TRUE) & ((4 + 1) != 5)
```

2. Evaluate !(3 + 5).  What does the ! operator do?  How is it different than !=?

```{r}
!(3 + 5)
```

Hint: ?`!`

The way in which ! is different than != is that != is strictly testing whether something is logically not equal to something else, while ! changes what logical statement something previously had.  In other words != is a way to examine existing logical conditions, while ! is a way to assign a different logical conditions by assigning previously true statements FALSE and previously false statements TRUE.


1.4 Missing Data

```{r, eval = FALSE}
swirl()
```

1.4.1 Practice

1. When you loaded the missing data swirl lesson, a vector called Nile should have been created in your environment.  How many missing values does this vector have?

```{r}
na_nile <- is.na(Nile)
sum(na_nile)
```

The Nile vector has zero missing values.

2. Wrap the sum() function around the is.na() function to find the answer to the above question in one step.  Hint: functions can be nested: f(g(x)).

```{r}
sum(is.na(Nile))
```

The Nile vector has zero missing values.

