---
title: "Lab 2"
output: html_document
---

1.1 Subsetting vectors using positional indicators and the bracket [] notation.

```{r, eval = FALSE}
library(swirl)
swirl()
```

1.1.1 Practice

1. Create a vector of 100 random normal values by typing rnorm(100) and assigning it to a vector.

```{r}
rand_hundred <- rnorm(100)
```

2. Print out the top 10 values.

```{r}
sorted <- sort(rand_hundred, decreasing = TRUE)
sorted[1:10]
```

3. Print out the values that are greater than 3.  Assign this to new vector object named y.

```{r}
y <- rand_hundred[rand_hundred > 3]
y
```


1.2 Subsetting Vectors Using Subset()

1. Read the help file for subset by typing ?subset.  Don't worry if it seems cryptic.

?subset


1.2.1 Practice

1. Repeat number 3 from the prior practice but use the subset() function. Assign this to a new vector object named z.

```{r}
z <- subset(rand_hundred, rand_hundred > 3)
```

2. Prove that these two methods produce the exact same subset by typing identical (y,z).

```{r}
identical (y,z)
```


1.3 Using Script Files

```{r}
add_for_me <- function(a,b){
  a + b
}

add_for_me(5,6)
```